---
title: "Covid-19 - Minimal report"
author: |
  | Human Bender
date: '`r Sys.Date()`'
output:
  html_document:
    #theme: flatly
    highlight: tango
    number_sections: true
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE, warning=FALSE, fig.align = "center")

##### Illustrating and Geo-mapping the SARS-CoV-2 #####
# Thomas Bury 2020
# 

# graphics, charts, colours and animation
require(ggplot2)
require(shadowtext)
require(viridis)
require(gghighlight)
require(paletteer)
require(ggmap)
require(scico)
require(gganimate)
require(gifski)
require(ggthemes)
require(cowplot)

# Data sources
require(nCov2019)
library(urbnmapr)
require(rnaturalearth)
require(maps)

# Data manipulation and simple features
require(dplyr)
require(sf)
require(tmap)
require(st)
require(data.table)
library(stringdist)

ClosestMatch2 = function(string, stringVector){
  stringVector[amatch(string, stringVector, maxDist=Inf)]
}

# Defining a temp ggplot theme complement
my_theme = theme(plot.title = element_text(color="gray50", size=14), legend.title = element_text(color="gray50"),
      legend.text = element_text(color = "gray50"))
```


```{r}
# Some helper functions
plot_top_states <- function(data, x.var, y.var, color.var, n.state, breaks, title, subtitle) {
  x.var = enquo(x.var)
  y.var = enquo(y.var)
  color.var = enquo(color.var)
  
  gg_plot = ggplot(data, aes(x = !! x.var, y = !! y.var, color = !! color.var)) +
    geom_line(size = 0.8) +
    geom_point(pch = 21, size = 1) +
    scale_y_continuous(expand = expansion(add = c(0,0.5)), breaks = breaks, labels = breaks) +
    #scale_y_log10(expand = expansion(add = c(0,0.5)), breaks = breaks, labels = breaks) +
    scale_x_continuous(expand = expansion(add = c(0,1.25))) +
    theme_fivethirtyeight() +
    labs(subtitle = subtitle, title = title) + 
    theme(legend.position = "none") +
    scale_color_paletteer_d("ggthemes::gdoc") +
    gghighlight(max(!! y.var), max_highlight = n.state, use_direct_label = T, 
                label_params = list(fill = NA, label.size = NA))
  return(gg_plot)
}


# facet plots for daily  and weekly new cases
plot_country_facet <- function(data, x.var, y.var, n.knots, title, n.col, n.row){
  y.var = enquo(y.var)
  x.var = enquo(x.var)
  
  gg_plot = ggplot(data = data, aes(x=!! x.var, y=!! y.var)) + 
    geom_point(size=3, color='gray20', alpha=0.5) +
    stat_smooth(se=FALSE, na.rm = TRUE, color = 'red3', size=1.25, 
                formula = y ~ s(x, bs = "cs", k = n.knots), method = 'gam') +
    #stat_smooth(se=FALSE, na.rm = TRUE, span=.5, size=1.5, color='blue') + 
    labs(title = title) + theme_fivethirtyeight() +
    theme(legend.position = "none") +
    scale_color_paletteer_d("ggthemes::gdoc") +
    facet_wrap(~country, scales = "free", ncol = n.col, nrow = n.row)
  return(gg_plot)
}

plot_last_week_map <- function(data, fill.var, min.val, max.val){
  fill.var = enquo(fill.var)
  
  cases_eu = ggplot(data = data, aes(fill=!! fill.var)) + geom_sf() + theme_map() +
    scale_fill_gradientn(colours = pals::ocean.thermal(100), trans = "log10", 
                         limits=c(min.val, max.val) )+
    ggtitle(label = 'Last observation') +
    my_theme
  return(cases_eu)
  
}

make_map_gif <- function(data, fill.var, group.var, map.data, title, name){
  fill.var = enquo(fill.var)
  group.var = enquo(group.var)
  
  cases_eu_anim = ggplot() + geom_sf(data = map.data) + 
    geom_sf(data = data, aes(fill=!! fill.var, group=!! group.var)) + 
    theme_map() + 
    scale_fill_gradientn(colours = pals::ocean.thermal(100), trans = "log10") +
    transition_time(week) + 
    labs(title = "Tot. confimed cases \n week:  {frame_time}") + 
    my_theme
  
  return(animate(cases_eu_anim, fps = 3, renderer = gifski_renderer()))
  anim_save(name)
}

plot_facet_map <- function(data, map.data, fill.var, group.var, title){
  fill.var = enquo(fill.var)
  group.var = enquo(group.var)
  
  map_facet = ggplot() + geom_sf(data = map.data) + 
    geom_sf(data = data, aes(fill=!! fill.var, group=!! group.var)) + 
    theme_map() +
    #scico::scale_fill_scico(palette = 'batlow', trans = "log10") + 
    scale_fill_gradientn(colours = pals::ocean.thermal(100), trans = "log10") +
    facet_wrap(~week) + 
    labs(title = title) + 
    my_theme
  return(map_facet)
}
```


# Loading the data

The data are collected from the Johns Hopkins University Center for Systems Science and Engineering and cleaned by datahub (see [This repo](https://github.com/datasets/covid-19)). In order to normalize the time axis, I filtered the data to retain the data since the 100th case has been reported. I filter out small countries (population less than 10000) and countries which are not in the world SF file, for convenience.

As the number of cases, deaths and healed are not measured/recorded each day, I need to expand the data set for each possible combination (country, time). If I don't do so, some countries will have missing records for some dates and the resulting charts and GIF will be "blinking". Hopefuly, tidyr makes that easy with the function complete (equivalent to expand and join) and with fill (filling NA with the previous non-NA value). I'll use log scales, so better to avoid zero values.



```{r}
d = fread('https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv')
death_per_country = fread("C:/Users/Thomas/R-open/sarscov2/deaths2018.csv")
```



```{r}
# aggregate by country 
d = d %>% mutate(Date = as.Date(date) ) %>%
  as_tibble  %>% 
  rename(time=Date, 
         country=location, 
         confirm=total_cases, 
         dead=total_deaths, 
         test = total_tests,
         confirm_pm=total_cases_per_million,
         dead_pm=total_deaths_per_million,
         test_pm=total_tests_per_thousand) %>% 
  group_by(country) %>%
  filter(confirm > 100 & country != "World") %>% #& country != "China"
  group_by(country) %>%
  mutate(days_since_100 = as.numeric(time - min(time))) %>%
  ungroup(country) %>% 
  tidyr::complete(time = seq.Date(min(time), max(time), by="day"))  %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, test, confirm_pm, dead_pm, test_pm) %>% ungroup(country)

# weekly aggregate
d_w = d %>% group_by(country, week = week(time)) %>% filter(time == max(time))
# weekly, only complete week
d_w = setDT(d_w)
d_w[, confirm_w := confirm - shift(confirm), by=c('country')]
d_w[d_w$confirm_w<0]$confirm_w = 0
country_top16 = d_w[week == max(week)-1][order(-confirm)][1:16, country]
d_w_subset = d_w[(country %in% country_top16) & (week < max(week))]
```



```{r}
# Find the last recorded 20 days of data
last_20_days = unique(d$time)
last_20_days = tail(last_20_days[order(last_20_days)], 20)
```



```{r}
# fetch natural earth data
world = rnaturalearth::countries110 %>% st_as_sf
world_countries = world %>% 
  filter(region_un %in% c('Africa', 'Americas', 'Asia', 'Europe', 'Oceania') & pop_est > 10000) %>% 
  select(name) %>% pull(name)

# Subset the world simple feature df
world_sf = world %>% 
  dplyr::filter(name %in% world_countries) %>% 
  select(name, geometry, region_un, pop_est)

# Europe simple features
# Load the map and filter the countries
suppressPackageStartupMessages(library(sf))
#world = st_as_sf(rnaturalearth::countries110)
europe = world_sf %>% dplyr::filter(region_un=="Europe" & name!='Russia') %>% select(name, geometry)
# A bounding box for continental Europe.
europe.bbox = st_polygon(list(matrix(c(-25,29,45,29,45,75,-25,75,-25,29),byrow = T,ncol = 2)))
europe.clipped = suppressWarnings(st_intersection(europe, st_sfc(europe.bbox, crs=st_crs(europe))))
eu_countries = europe.clipped %>% dplyr::distinct(name) %>% pull(name)
#all_eu_countries = world %>% filter(region_un=='Europe') %>% select(sovereignt) %>% st_set_geometry(NULL)

# Rem: as the number of cases, deaths and healed are not measured/recorded each day, 
# I need to expand the data set for each possible combination (country, time). 
# If I don't do so, some countries will have missing records for some dates
# and the resulting GIF will be "blinking". Hopefuly, tidyr makes that easy with the function
# complete (equivalent to expand and join) and with fill (filling NA with the previous non-NA value).
# I'll use log scales, so better to avoid zero values
world_df = d %>%
  as_tibble  %>% 
  filter(confirm >= 1 & country %in% world_countries) %>%  arrange(country, time) %>%
  tidyr::complete(country, time) %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, test, confirm_pm, dead_pm, test_pm) %>% ungroup %>% arrange(country, time)

world_df_dead = d %>% 
  as_tibble  %>%
  filter(dead >= 1) %>% arrange(country, time) %>%
  tidyr::complete(country, time) %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, test, confirm_pm, dead_pm, test_pm) %>% ungroup %>% arrange(country, time)

world_df_test = d %>% 
  as_tibble  %>% 
  filter(test_pm > 0) %>% arrange(country, time) %>%
  tidyr::complete(country, time) %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, test, confirm_pm, dead_pm, test_pm) %>% ungroup %>% arrange(country, time)

# weekly equivalent (less frames for gif, faster to compile)
world_w_df = d_w %>%
  as_tibble  %>% 
  filter(confirm >= 1 & country %in% world_countries) %>%  arrange(country, week) %>%
  tidyr::complete(country, week) %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, confirm_pm, dead_pm, test_pm) %>% ungroup %>% arrange(country, week)

world_w_df_dead = d_w %>%
  as_tibble  %>% 
  filter(dead >= 1) %>% arrange(country, week) %>%
  tidyr::complete(country, week) %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, confirm_pm, dead_pm, test_pm) %>% ungroup %>% arrange(country, week)

world_w_df_test = d_w %>%
  as_tibble  %>% 
  filter(test > 0) %>% arrange(country, week) %>%
  tidyr::complete(country, week) %>% group_by(country) %>% 
  tidyr::fill(confirm, dead, test, confirm_pm, dead_pm, test_pm) %>% ungroup %>% arrange(country, week)


# date of the first case in Europe
eu_date_fc = world_df %>% dplyr::filter(country %in% eu_countries) %>% group_by(country) %>% 
  slice(which.min(confirm)) %>% ungroup %>% select(time) %>% pull(time) %>% min %>% as.Date()
```


# How large is the pandemic

First of all, a sensible question to ask is what is the magnitude order of the pandemic. Because, there is a very sensitive trade-off between impacting the whole population of a country, at the risk that some households would go bankrupt and all the consequences that it would have, and save lives. 
Even if life has no value, induced poverty could affect a much more important part of the population and have long term consequences which could be even worst than the immediate effect of the virus.

Moreover, by digging deeper, the consequences of the virus could have been mitigated. Taking Belgium as an example, where 70% of the deaths are Patients from rest houses.

Note that countries are not reporting the same way, the definition of "death due to covid" varies across Europe (and worldwide) so numbers aren't straightforward to compare. Take BE and NL for instance, two neighboring countries with similar demographic characteristics, the percentage is twice larger in BE than in NL. It is partly due to the definition of "death due to covid".

The death rate per capita in Sweden is much lower while they didn't lock-down the country. Of course, there are a lot of factors, but we might ask: did the politics act rationally, and did they protect the population correctly while preserving the well being of the rest of the population, especially those struggling for the day to day living?

The best will be to compare the total number of deaths in 2020 compared to 2018 (year without "special event"). The 2020 number of deaths excess compared to 2018 will be due almost entierly to covid-19 (if there is no other severe event). If the reported excess mortality at the end of 2020 does not match the covid-19 excess w.r.t to a year without special event, then the counting was actually biased (the number of deaths being quite stable in occidental Europe, save under exceptional circumstances)


```{r}
country_list = death_per_country$country
last_df = world_w_df %>% group_by(country) %>% slice(which.max(week)) %>% dplyr::filter(country %in% country_list)
last_df = left_join(last_df, death_per_country, by = c("country" = "country") )
last_df = last_df %>% mutate(death_frac = dead/Deaths_2018)
last_df = last_df %>% tidyr::replace_na(list(death_frac = 0))
```

```{r, fig.height=7, fig.width=7}
ggplot(data = last_df, aes(x=reorder(country, death_frac), y=100*death_frac, color=death_frac)) + 
  geom_point(size = 2.5) + theme_fivethirtyeight() +  
  coord_flip() +
  scale_color_gradientn(colours = pals::kovesi.rainbow(100), guide = FALSE) +
  labs(subtitle = 'Percentage w.r.t the total number of deaths in 2018 (no special event).\nCounting not consistent accross countries', title = 'Covid-19 mortality (so far)') 
```


# Top 5 countries, aggregate figures - Worldwide and in EU

Top 5 countries, absolute and per million of inhabitants figures
Absolute numbers illustrate that the larger the country the more cases (not necessarily following the population ranking).
The relative numbers (per million of inhabitants) illustrate how the country is globally dealing with the crisis (of course, density and other parameters make the fight harder, but mostly the policy, how quick we react and analytics usage are involved).
BE is part of the 5 worst countries, using relative numbers. Especially for the number of deaths per million. This is mainly due to inefficient policies, especially regarding the rest houses (accounting for 50% of the deaths in BE)

## Worldwide

```{r, fig.width=15}
dead = plot_top_states(data = d, 
                       x.var=days_since_100,
                       y.var = dead, 
                       color.var=country,
                       n.state = 5L,
                       breaks = c(1000, 10000, 50000, 100000, 200000, 500000, 1000000),
                       title = "Nbr of deaths", 
                       subtitle = "Total deaths vs Number of days since 100th case")

tested = plot_top_states(data = d, 
                        x.var=days_since_100,
                        y.var = test_pm, 
                        color.var=country,
                        n.state = 5L,
                        breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 200, 500, 1000),
                        title = "Nbr of total tests", 
                        subtitle = "Total tests vs Number of days since 100th case")

confirmed = plot_top_states(data = d, 
                            x.var=days_since_100,
                            y.var = confirm, 
                            color.var=country,
                            n.state = 5L,
                            breaks = c(1000, 10000, 50000, 100000, 500000, 1000000, 2000000, 3000000, 40000000),
                            title = "Nbr of cases", 
                            subtitle = "Total cases vs Number of days since 100th case")



dead_pm = plot_top_states(data = d, 
                          x.var=days_since_100,
                          y.var = dead_pm, 
                          color.var=country,
                          n.state = 5L,
                          breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                          title = "Nbr of deaths per million capita", 
                          subtitle = "since 100th case")

tested_pm = plot_top_states(data = d, 
                            x.var=days_since_100,
                            y.var = test_pm, 
                            color.var=country,
                            n.state = 5L,
                            breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                            title = "Nbr of tested per thousand capita", 
                            subtitle = "since 100th case")

confirmed_pm = plot_top_states(data = d, 
                               x.var=days_since_100,
                               y.var = confirm_pm, 
                               color.var=country,
                               n.state = 5L,
                               breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                               title = "Nbr of cases per million capita", 
                               subtitle = "since 100th case")

# Arrange them on a grid, or save them separately 
cowplot::plot_grid(dead, dead_pm, ncol = 2)
cowplot::plot_grid(tested, tested_pm, ncol = 2)
cowplot::plot_grid(confirmed, confirmed_pm, ncol = 2)
```


## Europe

Highlighting the most impacted countries in EU.

```{r, fig.width=15}
dead = plot_top_states(data = d %>% filter(country %in% eu_countries), 
                       x.var=days_since_100,
                       y.var = dead, 
                       color.var=country,
                       n.state = 5L,
                       breaks = c(1000, 10000, 50000, 100000, 200000, 500000, 1000000),
                       title = "Nbr of deaths", 
                       subtitle = "Total deaths vs Number of days since 100th case")

healed= plot_top_states(data = d %>% filter(country %in% eu_countries), 
                        x.var=days_since_100,
                        y.var = heal, 
                        color.var=country,
                        n.state = 5L,
                        breaks = c(1000, 10000, 50000, 100000, 500000, 1000000, 2000000),
                        title = "Nbr of healed", 
                        subtitle = "Total healed vs Number of days since 100th case")

confirmed = plot_top_states(data = d %>% filter(country %in% eu_countries), 
                            x.var=days_since_100,
                            y.var = confirm, 
                            color.var=country,
                            n.state = 5L,
                            breaks = c(1000, 10000, 50000, 100000, 500000, 1000000, 2000000, 3000000, 40000000),
                            title = "Nbr of cases", 
                            subtitle = "Total cases vs Number of days since 100th case")



dead_pm = plot_top_states(data = d %>% filter(country %in% eu_countries), 
                          x.var=days_since_100,
                          y.var = dead_pm, 
                          color.var=country,
                          n.state = 5L,
                          breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                          title = "Nbr of deaths per million capita", 
                          subtitle = "since 100th case")

healed_pm = plot_top_states(data = d %>% filter(country %in% eu_countries), 
                            x.var=days_since_100,
                            y.var = heal_pm, 
                            color.var=country,
                            n.state = 5L,
                            breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                            title = "Nbr of healed per million capita", 
                            subtitle = "since 100th case")

confirmed_pm = plot_top_states(data = d %>% filter(country %in% eu_countries), 
                               x.var=days_since_100,
                               y.var = confirm_pm, 
                               color.var=country,
                               n.state = 5L,
                               breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                               title = "Nbr of cases per million capita", 
                               subtitle = "since 100th case")

# Arrange them on a grid, or save them separately 
cowplot::plot_grid(dead, dead_pm, ncol = 2)
cowplot::plot_grid(healed, healed_pm, ncol = 2)
cowplot::plot_grid(confirmed, confirmed_pm, ncol = 2)
```




## Some countries without national lockdown

The case of  Sweden, for instance, is a bit particular since they didn't apply a national lockdown while most of the other EU countries did. Did it work? Apparently, this decision was pretty armful for the population.

```{r, fig.width=15}
no_ld_countries = c('Sweden', 'Hungary', 'Latvia')
dead = plot_top_states(data = d %>% filter(country %in% no_ld_countries), 
                       x.var=days_since_100,
                       y.var = dead, 
                       color.var=country,
                       n.state = 5L,
                       breaks = c(1000, 10000, 50000, 100000, 200000, 500000, 1000000),
                       title = "Nbr of deaths", 
                       subtitle = "Total deaths vs Number of days since 100th case")

healed= plot_top_states(data = d %>% filter(country %in% no_ld_countries), 
                        x.var=days_since_100,
                        y.var = heal, 
                        color.var=country,
                        n.state = 5L,
                        breaks = c(1000, 10000, 50000, 100000, 500000, 1000000, 2000000),
                        title = "Nbr of healed", 
                        subtitle = "Total healed vs Number of days since 100th case")

confirmed = plot_top_states(data = d %>% filter(country %in% no_ld_countries), 
                            x.var=days_since_100,
                            y.var = confirm, 
                            color.var=country,
                            n.state = 5L,
                            breaks = c(1000, 10000, 50000, 100000, 500000, 1000000, 2000000, 3000000, 40000000),
                            title = "Nbr of cases", 
                            subtitle = "Total cases vs Number of days since 100th case")



dead_pm = plot_top_states(data = d %>% filter(country %in% no_ld_countries), 
                          x.var=days_since_100,
                          y.var = dead_pm, 
                          color.var=country,
                          n.state = 5L,
                          breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                          title = "Nbr of deaths per million capita", 
                          subtitle = "since 100th case")

healed_pm = plot_top_states(data = d %>% filter(country %in% no_ld_countries), 
                            x.var=days_since_100,
                            y.var = heal_pm, 
                            color.var=country,
                            n.state = 5L,
                            breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                            title = "Nbr of healed per million capita", 
                            subtitle = "since 100th case")

confirmed_pm = plot_top_states(data = d %>% filter(country %in% no_ld_countries), 
                               x.var=days_since_100,
                               y.var = confirm_pm, 
                               color.var=country,
                               n.state = 5L,
                               breaks = c(10, 50, 100, 500, 1000, 5000, 10000, 50000),
                               title = "Nbr of cases per million capita", 
                               subtitle = "since 100th case")

# Arrange them on a grid, or save them separately 
cowplot::plot_grid(dead, dead_pm, ncol = 2)
cowplot::plot_grid(healed, healed_pm, ncol = 2)
cowplot::plot_grid(confirmed, confirmed_pm, ncol = 2)
```



# Weekly new cases

## Worldwide 16 top countries

```{r, fig.width=15, fig.height=15}
plot_country_facet(data=d_w_subset, x.var=week, y.var=confirm_w, n.knots=9, 
                   title="Weekly new cases since the 100th case", 
                   n.col=4, n.row=4)
```

## Europe - 16 countries randomly chosen

```{r, fig.width=15, fig.height=15}
some_countries = c("Austria", "Belgium", "Czech Rep.", "France", "Germany", "Ireland",
                   "Italy", "Luxembourg", "Netherlands", "Poland",
                   "Portugal", "Romania", "Spain", "Sweden", "Switzerland", "Ukraine","United Kingdom")
d_w = setDT(d_w)
d_w_eu = d_w[(country %in% some_countries) & (week < max(week))]

plot_country_facet(data=d_w_eu, x.var=week, y.var=confirm_w, n.knots=9, 
                   title="Weekly new cases since the 100th case", 
                   n.col=4, n.row=4)
```


# Geomapping

```{r}
### Filter the sarscov2 data 
last_df = world_w_df %>% group_by(country) %>% 
  slice(which.max(week)) %>% 
  dplyr::filter(country %in% eu_countries) %>%
  ungroup()
d_eu = world_w_df %>% dplyr::filter(country %in% eu_countries & time > eu_date_fc)
cov_eu_ts = left_join(d_eu, europe.clipped, by = c("country" = "name") )
cov_eu_ts = cov_eu_ts %>% arrange(country, time)
cov_eu_ts = st_as_sf(cov_eu_ts)
### Join with the geometry
cov_eu = left_join(last_df, europe.clipped, by = c("country" = "name") )
cov_eu = st_as_sf(cov_eu)

# # Deaths
# last_df = world_w_df_dead %>% group_by(country) %>% slice(which.max(time)) %>% dplyr::filter(country %in% eu_countries)
# d_eu = world_w_df_dead %>% dplyr::filter(country %in% eu_countries & time > eu_date_fc)
# cov_eu_ts = left_join(d_eu, europe.clipped, by = c("country" = "name") )
# cov_eu_ts = cov_eu_ts %>% arrange(country, time)
# cov_eu_ts = st_as_sf(cov_eu_ts)
# ### Join with the geometry
# cov_eu = left_join(last_df, europe.clipped, by = c("country" = "name") )
# cov_eu = st_as_sf(cov_eu)
# 
# 
# # healed
# ### Filter the sarscov2 data 
# last_df = world_w_df_heal %>% group_by(country) %>% slice(which.max(time)) %>% dplyr::filter(country %in% eu_countries)
# d_eu = world_df_heal %>% dplyr::filter(country %in% eu_countries & time > eu_date_fc)
# cov_eu_ts = left_join(d_eu, europe.clipped, by = c("country" = "name") )
# cov_eu_ts = cov_eu_ts %>% arrange(country, time)
# cov_eu_ts = st_as_sf(cov_eu_ts)
# ### Join with the geometry
# cov_eu = left_join(last_df, europe.clipped, by = c("country" = "name") )
# cov_eu = st_as_sf(cov_eu)
```

## EU 

### Observations of the last full week

```{r}
max_val = max(cov_eu$confirm, na.rm = TRUE)
plot_last_week_map(data = cov_eu, fill.var = confirm, min.val=1000, max.val=max_val)
```

### Animation by week since the 100th case

```{r}
make_map_gif(data=cov_eu_ts, fill.var=confirm, group.var = country, map.data = europe.clipped, 
             title = "Tot. confimed cases \n week:  {frame_time}",
             name="eu_cases.gif")
```

or if you don't like GIFs, using facet:

```{r, fig.width=15}
plot_facet_map(data=cov_eu_ts, map.data=europe.clipped, 
               fill.var=confirm, group.var=country, 
               title="Tot. confimed cases")
```

## World

```{r}
# Clipping for better rendering
world.bbox = st_polygon(list(
  matrix(c(-170,-55, 170, -55, 170, 85,-170, 85, -170, -55),byrow = T,ncol = 2)))
world_sf_clipped = suppressWarnings(st_intersection(world_sf, st_sfc(world.bbox, crs=st_crs(world))))

# date of the first case in Europe
eu_date_fc = world_df %>% dplyr::filter(country %in% eu_countries) %>% group_by(country) %>% 
  slice(which.min(confirm)) %>% ungroup %>% select(time) %>% pull(time) %>% min %>% as.Date()


d_world = world_w_df %>% dplyr::filter(country %in% world_countries & time > eu_date_fc)
cov_world_ts = left_join(d_world, world_sf, by = c("country" = "name") )
cov_world_ts = cov_world_ts %>% arrange(country, time)
cov_world_ts = st_as_sf(cov_world_ts)
### Join with the geometry
last_df = world_w_df %>% group_by(country) %>% slice(which.max(week)) %>% dplyr::filter(country %in% world_countries) %>% ungroup()
cov_world = left_join(last_df, world_sf, by = c("country" = "name") )
cov_world = st_as_sf(cov_world)
```

### Observations of the last full week

```{r}
max_val = max(cov_world$confirm, na.rm = TRUE)
plot_last_week_map(data = cov_world, fill.var = confirm, min.val=1000, max.val = max_val)
```

### Animation by week since the 100th case

```{r}
make_map_gif(data=cov_world_ts, fill.var=confirm, group.var = country, map.data = world_sf, 
             title = "Tot. confimed cases \n week:  {frame_time}",
             name="world_cases.gif")
```



# USA - regional data

The regional US data are fetch from the NY Times github [NY Times github](https://github.com/nytimes/covid-19-data)

```{r}
# fetch the data
myfile = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
us_cov = fread(myfile)
# data are already aggregated by state (cumulative numbers)
# filtering the mainland states
states_sf = get_urbn_map("states", sf = TRUE) %>% st_as_sf
states_sf$state_fips = as.integer(states_sf$state_fips) 
main_states = sort(unique(states_sf$state_fips))
us_cov = us_cov[fips %in% main_states]
us_cov = us_cov[, `:=`(true_date = lubridate::ymd(date), week = week(lubridate::ymd(date)))]

# creating the weekly table for the gif and facet
us_cov_w = unique(us_cov, by = c('week', 'state'), fromLast=TRUE)

# fill the gaps, because there might be missing measures for some states and some dates.
us_cov = us_cov %>% as_tibble %>%
  tidyr::complete(true_date = seq.Date(min(true_date), max(true_date), by="day"))  %>% group_by(state) %>% 
  tidyr::fill(cases, deaths) %>% ungroup

us_cov_w = us_cov_w %>% as_tibble %>%
  tidyr::complete(week = seq(min(week), max(week)))  %>% group_by(state) %>% 
  tidyr::fill(cases, deaths) %>% ungroup

# Joining the cov and special features
us_sp_data = left_join(as.tbl(us_cov_w), states_sf, by = c("fips" = "state_fips"))
st_geometry(us_sp_data) = us_sp_data$geometry
last_date = max(us_cov_w$true_date)

last_20_days = unique(us_cov$true_date)
last_20_days = tail(last_20_days[order(last_20_days)], 20)

# Last recorded data, for the static visualization
us_sp_last = us_sp_data %>% filter(date == last_date)
```

## Last observations

```{r}
max_val = max(us_sp_last$cases, na.rm = TRUE)
cases_us = plot_last_week_map(data=us_sp_last, fill.var=cases, min.val=1000, max.val=max_val)
max_val = max(us_sp_last$deaths, na.rm = TRUE)
deaths_us = plot_last_week_map(data=us_sp_last, fill.var=deaths, min.val=1, max.val=max_val)

cowplot::plot_grid(cases_us, deaths_us, ncol = 2)
```

## Evolution over time

Using GIF

```{r}
make_map_gif(data=us_sp_data, fill.var=cases, group.var = state, map.data = states_sf, 
             title = "Tot. confimed cases \n week:  {frame_time}",
             name="us_cases.gif")
```

or facet plot

```{r, fig.width=15}
plot_facet_map(data=us_sp_data, 
               map.data=states_sf, 
               fill.var=cases, group.var=state,
               title="Tot. confimed cases")
```


## Evolution per State

```{r}
# Plotting evolution overtime since the 10th case
d_us = us_cov %>% 
  filter(cases > 10) %>%
  group_by(state) %>%
  mutate(days_since_10 = as.numeric(true_date - min(true_date))) %>%
  ungroup %>% 
  tidyr::complete(true_date = seq.Date(min(true_date), max(true_date), by="day"))  %>% group_by(state) %>% 
  tidyr::fill(cases, deaths) %>% ungroup
```

```{r, fig.width=15}
confirmed = plot_top_states(data = d_us, 
                             x.var=days_since_10,
                             y.var = cases, 
                             color.var=state,
                             n.state = 7L,
                             breaks = c(1000, 10000, 50000, 100000, 200000, 300000, 400000, 500000),
                             title = "Nbr of confirmed (top 7)", 
                             subtitle = "Total confirmed vs Number of days since 10th case")

deaths = plot_top_states(data = d_us, 
                            x.var=days_since_10,
                            y.var = deaths, 
                            color.var=state,
                            n.state = 7L,
                            breaks = c(1000, 10000, 20000, 30000, 40000, 50000, 60000, 70000),
                            title = "Nbr of death (top 7)", 
                            subtitle = "Total deaths vs Number of days since 10th case")



plot_grid(confirmed, deaths)
```

